---
title: "Directed Acyclic Graphs (DAGs)"
subtitle: "Causal Paths, d-Separation, and the Backdoor Criterion"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    chalkboard: true
    smaller: true
    scrollable: true
    math: katex
    transition: fade
    width: 1200
    height: 750
    footer: "POL 682 | DAGs"
    code-fold: true
    code-summary: "Show code"
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(ggdag)
library(dagitty)
library(patchwork)

az_blue <- "#2166AC"
az_red  <- "#B2182B"
```

# Part I: Why DAGs? {background-color="white" style="color: black;"}

Pictures are helpful

## Motivation

Gauss-Markov asked for a single thing of our covariates: they must be uncorrelated with $\epsilon$.

. . .

But *why* might $\text{cov}(X, \epsilon) \neq 0$?

- An omitted confounder is also in the error term
- $Y$ feeds back into $X$ (simultaneity)
- $X$ is measured with error

. . .

DAGs let us see these failures as **paths** on a diagram rather than derivations on a page.

## What is a DAG?

A **Directed Acyclic Graph** is a sketch of causal relationships.

- **Nodes** are variables
- **Arrows** are causal effects (X causes Y)
- **"Acyclic"**: no variable causes itself, even indirectly
- Everything you assume is on the diagram; everything you *don't* draw is assumed absent

. . .

DAGs don't tell you the causal structure. You provide these. This is a useful tool for ensuring **correct specification**.

## Exogeneity, Vizualized

An **exogenous** variable has no arrow pointing *into* it from anything in the model.

```{r}
#| label: fig-exogenous
#| fig-cap: "Exogeneity: X has no incoming arrows from Y or ε."
#| echo: false
#| fig-width: 7
#| fig-height: 3.5

dagify(
  y ~ x + e,
  labels = c("x" = "X", "y" = "Y", "e" = "ε (error)"),
  coords = list(x = c(x = 0, y = 2, e = 1), y = c(x = 0, y = 0, e = 1))
) |>
  ggdag(text = FALSE, use_labels = "label") +
  theme_dag()
```

. . .

$X$ is predetermined; nothing in the equation causes it. That is what OLS needs.

# Part II: Reading a DAG {background-color="white" style="color: black;"}

How do they "work"?

## Paths

A **path** is any sequence of arrows connecting two variables, in either direction.

. . .

- **Causal (directed) path.** All arrows point one way: $X \to M \to Y$.
- **Non-causal (backdoor) path.** At least one arrow points "wrong": $X \leftarrow U \to Y$.

. . .

Both kinds of paths transmit **statistical association**. Only the first is a **causal effect**.

. . .

The whole game of DAG analysis is separating the two.

## Three Junctions

Every path is built from three intersection types:

```{r}
#| label: fig-junctions
#| fig-cap: "Chains and forks transmit association and can be blocked by conditioning. Colliders block association but are opened by conditioning."
#| echo: false
#| fig-width: 11
#| fig-height: 3.5

chain <- dagify(m ~ x, y ~ m,
  coords = list(x = c(x = 0, m = 1, y = 2), y = c(x = 0, m = 0, y = 0)))
fork  <- dagify(x ~ u, y ~ u,
  coords = list(x = c(x = 0, u = 1, y = 2), y = c(x = 0, u = 1, y = 0)))
coll  <- dagify(c ~ x + y,
  coords = list(x = c(x = 0, c = 1, y = 2), y = c(x = 1, c = 0, y = 1)))

(ggdag(chain) + theme_dag() + ggtitle("Chain: X → M → Y")) |
(ggdag(fork)  + theme_dag() + ggtitle("Fork: X ← U → Y"))  |
(ggdag(coll)  + theme_dag() + ggtitle("Collider: X → C ← Y"))
```

## How Conditioning Behaves

| Intersection | Shape | Open by default? | Conditioning on middle |
|----------|-------|------------------|------------------------|
| **Chain** | $X \to M \to Y$ | Yes | **Blocks** |
| **Fork** | $X \leftarrow U \to Y$ | Yes | **Blocks** |
| **Collider** | $X \to C \leftarrow Y$ | **No** | **Opens** |

. . .

**The surprise is the collider.** Controlling for a common effect *creates* an association between its parents. Things that *should be* independent become *conditionally* dependent.

## d-Separation

Two variables are **d-separated** by a set $\mathbf{Z}$ when *every* path between them is blocked.

. . .

A path is **blocked** by $\mathbf{Z}$ if:

- $\mathbf{Z}$ contains a non-collider on the path (a chain or fork middle), **or**
- The path contains a collider, and neither it nor its descendants are in $\mathbf{Z}$

. . .

d-separation $\Rightarrow$ conditional independence. This is the engine.

## The Backdoor Criterion

> Pick a set $\mathbf{Z}$ such that:
> **(a)** $\mathbf{Z}$ blocks every backdoor path from $X$ to $Y$,
> **(b)** $\mathbf{Z}$ contains no descendants of $X$.

. . .

If such a $\mathbf{Z}$ exists, regressing $Y$ on $X$ and $\mathbf{Z}$ recovers the **causal effect** of $X$ on $Y$.

. . .

This is the formal rule behind "controlling for" something. Not everything deserves to be a control.

# Part III: A Running Example {background-color="white" style="color: black;"}

College, knowledge, SES.

## The Question

Does **college** ($X$) cause **voter engagement** ($Y$)?

. . .

Two other variables are in play:

- **Parental SES** ($U$): affects both whether you go to college and whether you vote.
- **Political knowledge** ($M$): college raises knowledge; knowledge raises engagement.

## The DAG

```{r}
#| label: fig-backdoor
#| fig-cap: "College (X) → Engagement (Y), both directly and through political knowledge (M). Parental SES (U) is a confounder."
#| echo: false
#| fig-width: 8
#| fig-height: 4

dagify(
  y ~ x + m + u,
  m ~ x,
  x ~ u,
  labels = c("x" = "College (X)",
             "m" = "Knowledge (M)",
             "y" = "Engagement (Y)",
             "u" = "Parental SES (U)"),
  coords = list(x = c(x = 0, m = 1, y = 2, u = 0.5),
                y = c(x = 0, m = 0.6, y = 0, u = 1))
) |>
  ggdag(text = FALSE, use_labels = "label") +
  theme_dag()
```

## Listing the Paths

From $X$ to $Y$:

1. $X \to Y$ — **causal** (direct).
2. $X \to M \to Y$ — **causal** (indirect, through the mediator).
3. $X \leftarrow U \to Y$ — **backdoor** (non-causal; the source of bias).

. . .

We want the **total effect** of $X$ on $Y$: close path 3, leave paths 1 and 2 alone.

## Applying the Criterion

- **Condition on $U$?** Yes — $U$ is the fork middle on path 3; adjusting blocks it. $U$ is not a descendant of $X$. Allowed.
- **Condition on $M$?** No — $M$ is a chain middle on path 2, which is **causal**. Closing it throws away the indirect effect.

. . .

**Valid adjustment set:** $\mathbf{Z} = \{U\}$.

. . .

- `lm(Y ~ X + U)` → **total** causal effect.
- `lm(Y ~ X + U + M)` → **direct** effect only.
- `lm(Y ~ X)` → biased by the $X \leftarrow U \to Y$ backdoor.

## Verify with `dagitty`

```{r}
#| label: adj-sets-example
#| echo: true

g_ex <- dagitty::dagitty("dag {
  U -> X
  U -> Y
  X -> M
  M -> Y
  X -> Y
}")
dagitty::adjustmentSets(g_ex, exposure = "X", outcome = "Y")
```

. . .

`dagitty` returns every minimal adjustment set that blocks all backdoors.

# Part IV: Colliders and "Bad Controls" {background-color="white" style="color: black;"}

Why more controls is not always better.

## The Collider 

A collider has two arrows pointing *in*: $X \to C \leftarrow Y$.

. . .

- **Unadjusted:** no association flows through $C$.
- **Adjusted:** an association appears between $X$ and $Y$ that is not causal.

. . .

This is how "controlling for everything" creates bias that wasn't there.

## A Classic Example

Suppose $X$ = athletic talent, $Y$ = academic talent, and $C$ = admission to an elite college (which demands either).

. . .

In the general population, $X$ and $Y$ may be independent. Among **admitted students only** (conditioning on $C$), they appear **negatively correlated** — if you got in without academic talent, you likely had athletic talent.

. . .

The negative association is an artifact of conditioning on the collider.

## Three Rules for Specification

From the backdoor criterion, three operational rules:

1. **Block every backdoor.** Control for confounders (forks).
2. **Do not condition on a collider.** Including one creates spurious association between its parents.
3. **Do not condition on a post-treatment mediator** if you want the total effect. Chains carry part of the causal signal; closing them throws it away.

. . .

`adjustmentSets()` enforces all three automatically.

## When No Valid Set Exists

```{r}
#| label: adj-sets-unobs
#| echo: true

g <- dagitty::dagitty("dag {
  U [unobserved]
  Z -> X
  X -> Y
  U -> X
  U -> Y
}")
dagitty::adjustmentSets(g, exposure = "X", outcome = "Y")
```

. . .

Empty. $U$ is unobserved, so no observable set blocks the backdoor through it.

. . .

**This is exactly where instrumental variables enter** — see the exogeneity lecture.

# Part V: Why This Matters for OLS {background-color="white" style="color: black;"}

The DAG as a diagnosis.

## DAGs and Gauss-Markov

Gauss-Markov demanded $\text{cov}(X, \epsilon) = 0$.

. . .

A DAG makes this vivid: the error term represents **everything we haven't drawn**. If an unobserved $U$ is a fork between $X$ and $Y$, then $U$ lives in $\epsilon$ and is correlated with $X$.

. . .

**OLS is unbiased iff there is no open backdoor path from $X$ to $Y$.**

## What Can Go Wrong

Three routes to a correlated error term:

- **Omitted confounder** (fork): $X \leftarrow U \to Y$.
- **Simultaneity**: $Y \to X$ alongside $X \to Y$.
- **Measurement error in X**: observed $X$ depends on true $X^*$ and a noise $\gamma$; $\gamma$ is in the error.

. . .

All three reduce to the same statement: **there is an open backdoor from $X$ to the outcome's error.**

## Key Takeaways

- A DAG encodes your causal claims so they can be audited.
- Chains and forks transmit association; conditioning **blocks** them.
- Colliders block by default; conditioning **opens** them.
- **Backdoor criterion**: block every non-causal path without touching any causal path.
- "Control for everything" is wrong — some variables are colliders or mediators.
- When no adjustment set exists, instrumental variables are the natural next step.
