---
title: "Exogeneity, Endogeneity, and Instrumental Variables"
subtitle: "2SLS, Weak Instruments, and a Worked WSS 2020 Example"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    chalkboard: true
    smaller: true
    scrollable: true
    math: katex
    transition: fade
    width: 1200
    height: 750
    footer: "POL 682 | Exogeneity & IV"
    code-fold: true
    code-summary: "Show code"
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(ggdag)
library(dagitty)
library(patchwork)

az_blue <- "#2166AC"
az_red  <- "#B2182B"
```

# Part I: Exogeneity {background-color="white" style="color: black;"}

The assumption that pulls the rest together.

## Definition

**Exogeneity** of $X$: $X$ is determined outside the model; it is uncorrelated with the error term.

. . .

In the DAG: no arrow points into $X$ from anything in the model.

. . .

In algebra: $\text{cov}(X, \epsilon) = 0$.

. . .

**Endogenous** is the opposite — $X$ is determined inside the model, and the covariance with $\epsilon$ is non-zero.

## Why Exogeneity is the Umbrella Assumption

Several Gauss-Markov conditions collapse into exogeneity:

- **Proper specification**: omitted confounders get absorbed into $\epsilon$ and correlate with $X$.
- **Measurement error in X**: the noise becomes part of the composite error.
- **Simultaneity**: $Y$ feeds back into $X$, so $X$ reflects $\epsilon$.

. . .

All three violations say the same thing: **there is an open backdoor from $X$ to the error.**

## Three Routes to Endogeneity

```{r}
#| label: fig-endog-types
#| fig-cap: "Three ways exogeneity breaks: an omitted confounder, simultaneity, and measurement error in X."
#| echo: false
#| fig-width: 11
#| fig-height: 3.8

d1 <- dagify(
  y ~ x + u, x ~ u,
  labels = c("x" = "X", "y" = "Y", "u" = "U (unobs)"),
  coords = list(x = c(x = 0, y = 2, u = 1), y = c(x = 0, y = 0, u = 1))
)
d2 <- dagify(
  y ~ x, x ~ y,
  labels = c("x" = "X", "y" = "Y"),
  coords = list(x = c(x = 0, y = 2), y = c(x = 0, y = 0))
)
d3 <- dagify(
  y ~ xstar, xobs ~ xstar + gamma,
  labels = c("xstar" = "X*", "xobs" = "X (obs)",
             "y" = "Y", "gamma" = "γ"),
  coords = list(x = c(xstar = 0, xobs = 1, gamma = 1, y = 2),
                y = c(xstar = 0, xobs = 1, gamma = 2, y = 0))
)

(ggdag(d1, text = FALSE, use_labels = "label") + theme_dag() + ggtitle("Omitted confounder")) |
(ggdag(d2, text = FALSE, use_labels = "label") + theme_dag() + ggtitle("Simultaneity")) |
(ggdag(d3, text = FALSE, use_labels = "label") + theme_dag() + ggtitle("Measurement error"))
```

. . .

In each case OLS sees only the muddled association. The coefficient is biased.

# Part II: Instrumental Variables {background-color="white" style="color: black;"}

When adjustment sets run out.

## The Problem

If the confounder $U$ is **unobserved**, no adjustment set blocks the backdoor.

. . .

`adjustmentSets()` returns empty. OLS is stuck.

. . .

Enter the **instrumental variable**: an exogenous nudge on $X$ that cannot reach $Y$ except through $X$.

## A Valid Instrument

```{r}
#| label: fig-instrument
#| fig-cap: "A valid instrument Z: relevance (Z → X) and exclusion (no direct Z → Y and no Z ↔ U)."
#| echo: false
#| fig-width: 8
#| fig-height: 4

dagify(
  y ~ x + u,
  x ~ z + u,
  labels = c("z" = "Z (instrument)", "x" = "X", "y" = "Y",
             "u" = "U (unobserved)"),
  coords = list(x = c(z = -1, x = 0, y = 2, u = 1),
                y = c(z = 0, x = 0, y = 0, u = 1))
) |>
  ggdag(text = FALSE, use_labels = "label") +
  theme_dag()
```

## Two Conditions

1. **Relevance.** $Z \to X$. The instrument actually moves $X$.
2. **Exclusion.** $Z$ does not reach $Y$ except through $X$. No direct $Z \to Y$. No link to the confounder $U$.

. . .

Relevance is **testable** (first-stage F). Exclusion is **not** — you argue it on theory.

. . .

If both hold, the variation in $X$ induced by $Z$ is **exogenous**. That variation is clean.

## Two-Stage Least Squares

$$
y_i = b_0 + b_1 x_i + e_{1i}, \qquad x_i = a_0 + a_1 z_i + e_{2i}.
$$

. . .

**Stage 1.** Regress $X$ on $Z$ (and other exogenous controls). Keep $\hat{x}$.

**Stage 2.** Regress $Y$ on $\hat{x}$.

. . .

$\hat{x}$ is the part of $X$ driven by the instrument — the clean variation. Using it in stage 2 purges the correlation with $U$.

. . .

Do not actually run two separate `lm()` calls — the standard errors will be wrong. Use `AER::ivreg()`.

# Part III: Identification {background-color="white" style="color: black;"}

When can 2SLS even be estimated?

## The Order Condition

Outcome equation with endogenous $x_2$ and exogenous controls $z_1, z_2$:

$$y = \beta_1 x_2 + \gamma_1 z_1 + \gamma_2 z_2 + \epsilon_1.$$

. . .

If the first stage uses only $z_1, z_2$, then

$$\hat{x}_2 = \hat\pi_1 z_1 + \hat\pi_2 z_2$$

is a linear combination of variables **already in the outcome equation**. $X^T X$ is singular. **Not identified.**

## The Fix

Add an **excluded instrument** $z_3$ that appears in the first stage but **not** the outcome equation:

$$x_2 = \pi_1 z_1 + \pi_2 z_2 + \pi_3 z_3 + \epsilon_2.$$

. . .

Now $\hat{x}_2$ contains a piece ($\hat\pi_3 z_3$) that is not spanned by the second-stage regressors.

. . .

**Order condition:** at least one instrument must appear in the first stage but not the second. One per endogenous regressor.

## 2SLS in Matrix Form

Stack the regressors into $\hat{W} = [\,\hat{X}\ \ Z\,]$, where

- $\hat{X}$ = fitted values of endogenous regressors from the first stage
- $Z$ = included exogenous regressors (intercept, controls, dummies)

. . .

$$\hat{\boldsymbol{\beta}}_{2SLS} = (\hat{W}^T \hat{W})^{-1} \hat{W}^T \mathbf{y}.$$

. . .

Variance is $\sigma^2 (\hat{W}^T \hat{W})^{-1}$, where $\sigma^2$ is computed from residuals using the **original** $X$, not $\hat{X}$. Manual two-step regressions miss this correction; `ivreg()` does it automatically.

# Part IV: Testing for Endogeneity {background-color="white" style="color: black;"}

Is 2SLS even needed?

## The Tradeoff

Under exogeneity:

- **OLS**: unbiased, consistent, efficient (minimum-variance among linear unbiased estimators)
- **2SLS**: consistent but **biased in finite samples**, and always less efficient (larger SEs)

. . .

So if $X$ really is exogenous, 2SLS costs precision for no gain. If $X$ is endogenous, OLS is wrong and 2SLS is necessary.

. . .

We want a test.

## Unbiasedness vs. Consistency

A distinction that matters here:

- **Unbiased**: $E[\hat\beta] = \beta$ at every sample size.
- **Consistent**: $\hat\beta \xrightarrow{p} \beta$ as $n \to \infty$.

. . .

OLS (under Gauss-Markov) is **both**.

**2SLS** is **consistent but biased in finite samples**. First-stage noise correlates with second-stage error; bias shrinks as $n$ grows but is non-trivial in small samples.

## The Residual Test

Save first-stage residuals $\hat{\delta}$ and include them in the outcome equation:

$$y = \beta_1 x_1 + \beta_2 \hat{\delta} + \gamma_1 z_1 + \gamma_2 z_2 + \epsilon.$$

. . .

If $\hat\beta_2 \neq 0$, the part of $x_1$ not explained by the instruments is correlated with the outcome's error — i.e., $x_1$ is endogenous.

## The Hausman Test

Compare the two coefficient vectors directly:

$$H = (\hat\beta_c - \hat\beta_e)^T (V_c - V_e)^{-1} (\hat\beta_c - \hat\beta_e),$$

distributed $\chi^2$ with df equal to the number of compared coefficients.

. . .

- **Small $H$**: OLS and 2SLS agree. Prefer the efficient OLS.
- **Large $H$**: They disagree. Prefer 2SLS.

. . .

`ivreg(..., diagnostics = TRUE)` reports the Wu-Hausman version by default.

## Weak Instruments

A valid instrument may still be **weak** — $Z$ explains very little variation in $X$.

. . .

Rough diagnostic: first-stage $F$ on the excluded instruments. A rule of thumb is $F > 10$; more recent work pushes that threshold higher.

. . .

**Weak instruments are bad even when valid**: 2SLS estimates become noisy, finite-sample bias grows, and $\hat{x}$ collapses back into the other regressors (collinearity in the second stage).

# Part V: Worked Example — WSS 2020 {background-color="white" style="color: black;"}

A randomized instrument in the wild.

## The Question

**Does endorsing political protest affect trust in government?**

. . .

People who endorse protest differ on many unobservables — prior distrust, political engagement, ideology. `attend_march` is almost surely **endogenous**.

## The Randomized Instrument

The WSS 2020 randomized respondents to one of two versions of the protest question:

- **Control**: standard "would you attend a march or rally"
- **Treatment**: the same plus "even if it turns violent"

. . .

Treated respondents endorse attending a march at **lower rates** — relevance.

. . .

Assignment is random — exogeneity of $Z$ is by **design**, not by argument.

## The Assumed DAG

```{r}
#| label: fig-wss-dag
#| fig-cap: "Randomized Z → X → Y. U confounds X–Y; randomization ensures no arrow into Z."
#| echo: false
#| fig-width: 8
#| fig-height: 4

dagify(
  y ~ x + u,
  x ~ z + u,
  labels = c(
    "z" = "Violent prompt (Z)",
    "x" = "Attend-march (X)",
    "y" = "Trust gov't (Y)",
    "u" = "U (unobserved)"
  ),
  coords = list(x = c(z = -1, x = 0, y = 2, u = 1),
                y = c(z = 0, x = 0, y = 0, u = 1))
) |>
  ggdag(text = FALSE, use_labels = "label") +
  theme_dag()
```

. . .

**Exclusion restriction**: the wording manipulation affects trust *only* through its effect on stated march endorsement.

## Step 1: Naive OLS

```r
ols_fit <- lm(trustGovernment ~ attend_march, data = d)
summary(ols_fit)$coefficients
```

. . .

This is what we'd report if we ignored the endogeneity problem. It mixes the causal effect with whatever is in $U$.

## Step 2: The First Stage

```r
first_stage <- lm(attend_march ~ violent_treat, data = d)
summary(first_stage)$coefficients
first_stage_F <- summary(first_stage)$fstatistic[1]
first_stage_F
```

. . .

Checks **relevance**: does the randomized prompt move `attend_march`? The coefficient and the F-statistic both speak to this.

## Step 3: 2SLS via `ivreg`

```r
library(AER)
iv_fit <- ivreg(trustGovernment ~ attend_march | violent_treat, data = d)
summary(iv_fit, diagnostics = TRUE)$coefficients
summary(iv_fit, diagnostics = TRUE)$diagnostics
```

. . .

The `diagnostics` block reports three tests:

1. **Weak instruments F** — formal first-stage test
2. **Wu-Hausman** — is 2SLS telling you something OLS missed?
3. **Sargan** — only computed with *over*-identification (more instruments than endogenous regressors)

## Step 4: Compare

```r
data.frame(
  estimator = c("OLS", "2SLS"),
  estimate  = c(coef(ols_fit)["attend_march"],
                coef(iv_fit) ["attend_march"]),
  se        = c(summary(ols_fit)$coefficients["attend_march", "Std. Error"],
                summary(iv_fit )$coefficients["attend_march", "Std. Error"])
)
```

. . .

Three things almost always happen:

- **The 2SLS point estimate differs from OLS** (sometimes dramatically)
- **The 2SLS standard error is larger** (IV discards information)
- **Weak instruments punish you twice** — noisy *and* biased in finite samples

# Part VI: Systems of Equations {background-color="white" style="color: black;"}

When endogeneity becomes structural.

## Simultaneity

Two variables that move together:

$$
x_2 = b_{00} + b_{01} x_1 + e_1
$$

$$
x_1 = b_{10} + b_{11} x_2 + e_2
$$

. . .

Both endogenous. $\text{cov}(x_1, e_1) \neq 0$ and $\text{cov}(x_2, e_2) \neq 0$. OLS biased on both.

. . .

The system is **not identified** without more information.

## Identification via Exclusion

Add an exogenous variable **unique to each equation**:

$$
x_2 = b_{00} + b_{01} x_1 + \gamma_1 z_1 + e_1, \qquad
x_1 = b_{10} + b_{11} x_2 + \gamma_2 z_2 + e_2.
$$

. . .

- $z_1$ predicts $x_1$ but is excluded from the $x_2$ equation — identifies the $x_2$ equation.
- $z_2$ does the same, symmetrically.

. . .

**Rule**: each equation in a simultaneous system needs at least one instrument excluded from it but present in the equation that generates the endogenous RHS variable.

## Recursive vs. Non-Recursive

```{r}
#| label: fig-rec-nonrec
#| fig-cap: "Recursive (left): ordered, no feedback. Non-recursive (right): X₁ ↔ X₂ — OLS biased; instruments required."
#| echo: false
#| fig-width: 10
#| fig-height: 3.5

rec <- dagify(
  x2 ~ x1,
  x3 ~ x2 + x1,
  labels = c(x1 = "X\u2081", x2 = "X\u2082", x3 = "X\u2083"),
  coords = list(x = c(x1 = 0, x2 = 1, x3 = 2),
                y = c(x1 = 0, x2 = 0, x3 = 0))
)
nonrec <- dagify(
  x1 ~ x2,
  x2 ~ x1,
  labels = c(x1 = "X\u2081", x2 = "X\u2082"),
  coords = list(x = c(x1 = 0, x2 = 1),
                y = c(x1 = 0, x2 = 0))
)

(ggdag(rec,    text = FALSE, use_labels = "label") + theme_dag() + ggtitle("Recursive")) |
(ggdag(nonrec, text = FALSE, use_labels = "label") + theme_dag() + ggtitle("Non-recursive"))
```

. . .

- **Recursive, uncorrelated errors**: OLS equation-by-equation is consistent.
- **Non-recursive**: cycle in the DAG. OLS biased on every equation in the cycle. Needs instruments.

## Direct vs. Indirect Effects

Substituting the $x_1$ equation into the $x_2$ equation:

$$
x_2 = \omega^{*} z_1 + \omega_3 z_2 + e^{*}_1,
\qquad
\omega^{*} = \omega_2 + b_{11}\omega_1.
$$

. . .

$\omega^{*}$ is the **total effect** of $z_1$ on $x_2$: direct ($\omega_2$) plus indirect through $x_1$ ($b_{11}\omega_1$).

. . .

This is specification bias from POL 682, now turned into a tool: a correctly-specified recursive system lets us **decompose** total effects.

# Part VII: Practical Considerations {background-color="white" style="color: black;"}

Why IV is harder than it looks.

## Finding a Valid Instrument Is Hard

- Must move $X$ (relevance)
- Must *not* reach $Y$ except through $X$ (exclusion)
- Must not relate to the unobserved confounder

. . .

Most candidate instruments fail at least one check. Randomization, natural experiments, and institutional quirks are the usual suspects.

## When Instruments Fail

- **Invalid exclusion** → 2SLS biased, sometimes worse than OLS
- **Weak first stage** → noisy 2SLS and finite-sample bias toward OLS
- **Over-identified with conflicting instruments** → Sargan test should catch disagreement among instruments, but only if at least one is actually valid

## `ivreg` Syntax Recap

```r
library(AER)
fit <- ivreg(y ~ x1 + x2 | z1 + x2, data = df)
```

. . .

- **Before `|`**: the outcome equation (endogenous + exogenous controls).
- **After `|`**: the *full* set of exogenous regressors (instruments **and** included controls that are exogenous).

. . .

Exogenous controls appear on **both** sides of the `|`.

## Key Takeaways

- **Exogeneity** is the umbrella: specification, measurement, and simultaneity all collapse into it.
- With no valid adjustment set, **instrumental variables** are the tool.
- A valid IV needs **relevance** (testable) and **exclusion** (argued).
- **2SLS** isolates the instrument-driven variation in $X$ and uses that to estimate the effect.
- **2SLS is biased in finite samples** (consistent only) and less efficient than OLS — use it when you must, not by default.
- **Weak instruments** make things worse on both fronts.
- In systems of equations, each equation needs at least one **excluded** instrument for identification.
