---
title: "Measurement Error"
subtitle: "Attenuation Bias, Reliability, and Corrections"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    chalkboard: true
    smaller: true
    scrollable: true
    math: katex
    transition: fade
    width: 1200
    height: 750
    footer: "POL 682 | Measurement Error"
    code-fold: true
    code-summary: "Show code"
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)

az_blue <- "#2166AC"
az_red  <- "#B2182B"
```

# Part I: Measurement Error {background-color="white" style="color: black;"}

What happens when we can't observe the variables we care about?

## Roadmap

Every variable we measure has error. The question is whether that error biases our estimates.

. . .

- Measurement error in the **dependent variable**: OLS is still BLUE, but standard errors inflate
- Measurement error in the **independent variable**: OLS is biased — **attenuation bias**
- **Reliability** gives a formal handle on how bad it is
- Corrections

. . .

**The asymmetry matters.** Noisy X and noisy Y are not the same problem.

## Two Kinds of Error

Let the true data-generating process be

$$Y^{*}_i = \alpha + \beta X^{*}_i + \epsilon_i.$$

. . .

We may only observe **noisy versions**:

$$Y_i = Y^{*}_i + u_i \qquad \text{(error in DV)}$$

$$X_i = X^{*}_i + \gamma_i \qquad \text{(error in IV)}$$

. . .

Classical measurement error assumes $u_i$ and $\gamma_i$ are mean-zero, uncorrelated with the true values and with $\epsilon_i$.

# Part II: Error in the Dependent Variable {background-color="white" style="color: black;"}

The easy case.

## Substituting Noisy Y

Starting with the true model and substituting the observed $Y_i$,

$$Y_i - u_i = \alpha + \beta X_i + \epsilon_i$$

$$Y_i = \alpha + \beta X_i + (\epsilon_i + u_i).$$

. . .

Two error components now, but both are uncorrelated with $X_i$.

. . .

**OLS is still BLUE.** Unbiased. Consistent. Just noisier.

## What Happens to the Variance

The variance of the slope estimator becomes

$$\text{var}(b) = \frac{\sigma^2_u + \sigma^2_{\epsilon}}{\sum (X_i - \bar{X})^2}.$$

. . .

- Numerator grows with $\sigma^2_u$ — more noise in Y, bigger standard errors
- Expected value of $b$ is unchanged: $E[b] = \beta$
- Inference is conservative, not biased

## Simulation: Error in Y

```{r}
#| label: me-dv-sim
#| echo: true

set.seed(2026)
n <- 500
x <- rnorm(n, 0, 1)
y_star <- 0 + 1 * x + rnorm(n, 0, 1)

sigma_u_grid <- c(0, 0.5, 1, 2, 4)

me_dv <- lapply(sigma_u_grid, function(su) {
  y_obs <- y_star + rnorm(n, 0, su)
  fit   <- lm(y_obs ~ x)
  data.frame(
    sigma_u  = su,
    beta_hat = coef(fit)[2],
    se_beta  = sqrt(diag(vcov(fit)))[2]
  )
}) |> do.call(rbind, args = _)

me_dv
```

. . .

Coefficients hover around 1; standard errors grow roughly with $\sigma_u$.

# Part III: Error in the Independent Variable {background-color="white" style="color: black;"}

The hard case.

## Substituting Noisy X

Start with $Y_i = \alpha + \beta X^{*}_i + \epsilon_i$ and substitute $X^{*}_i = X_i - \gamma_i$:

$$Y_i = \alpha + \beta X_i + \epsilon_i - \beta \gamma_i.$$

. . .

Combined error: $z_i = \epsilon_i - \beta \gamma_i$.

. . .

And now:

$$\text{cov}(X, z) = \text{cov}(X^{*} + \gamma,\ \epsilon - \beta\gamma) = -\beta \, \sigma^2_{\gamma}.$$

. . .

The regressor and the error are **correlated**. Gauss-Markov broken. OLS biased.

## Attenuation Bias

$$\text{plim} \, \hat{\beta} = \beta \cdot \underbrace{\frac{\sigma^2_{X^{*}}}{\sigma^2_{X^{*}} + \sigma^2_{\gamma}}}_{\text{reliability}}.$$

. . .

- The ratio is the **reliability** of $X$, always in $[0,1]$
- $\sigma^2_\gamma \to 0$: reliability $\to 1$, $\hat\beta \to \beta$
- $\sigma^2_\gamma$ grows: reliability shrinks, $\hat\beta$ **shrinks toward zero**

. . .

**Attenuation bias.** Noisy predictors always look "flatter" than they really are.

## Simulation

```{r}
#| label: fig-attenuation
#| fig-cap: "As measurement error in X grows, the OLS slope shrinks toward zero. Dashed line is the true β = 1."
#| echo: true
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 4.5

set.seed(2026)
n <- 1000
beta_true <- 1.0

x_star  <- rnorm(n, 0, 1)
epsilon <- rnorm(n, 0, 1)
y       <- 0 + beta_true * x_star + epsilon

sigma_gamma_grid <- seq(0, 2, by = 0.1)
results <- data.frame(sigma_gamma = sigma_gamma_grid, beta_hat = NA_real_)

for (i in seq_along(sigma_gamma_grid)) {
  sg <- sigma_gamma_grid[i]
  x_obs <- x_star + rnorm(n, 0, sg)
  results$beta_hat[i] <- coef(lm(y ~ x_obs))[2]
}

ggplot(results, aes(x = sigma_gamma, y = beta_hat)) +
  geom_hline(yintercept = beta_true, linetype = "dashed", color = az_red) +
  geom_line(color = az_blue, linewidth = 1) +
  geom_point(color = az_blue, size = 2) +
  labs(
    x = expression(sigma[gamma] ~ "(measurement error in X)"),
    y = expression(hat(beta))
  ) +
  theme_minimal(base_size = 14)
```

## Summary: DV vs. IV

| Error in | Bias? | Variance? | Fix? |
|----------|-------|-----------|------|
| Y (DV)   | No    | Inflated  | Accept the wider CIs |
| X (IV)   | **Yes** — toward 0 | Distorted | Correct or instrument |

. . .

The practical rule: **worry more about noisy predictors than noisy outcomes.**

# Part IV: Corrections {background-color="white" style="color: black;"}

If measurement is noisy, what can we do?

## Options

- **Factor analysis / SEM.** Model a latent construct separate from its noisy indicators.
- **Multi-item scales.** Averaging $k$ noisy indicators of the same construct reduces measurement-error variance roughly by $1/k$ (classical test theory).
- **Item Response Theory.** Model the mapping from latent trait to discrete responses.
- **Two-stage least squares (IV).** An instrument correlated with $X^{*}$ but not with $\gamma$ or $\epsilon$ recovers a consistent $\hat\beta$.

. . .

We'll see 2SLS in the exogeneity lecture — measurement error in X is a form of endogeneity.

## Practical Advice

- Measure constructs as accurately as possible before you worry about estimators
- Multi-item measures are cheap; they often dominate post-hoc corrections
- Check for coding errors. Always.
- If attenuation is plausible, report that $\hat\beta$ is a **lower bound** on $|\beta|$

. . .


## Key Takeaways

- Classical measurement error in **Y**: OLS still BLUE; standard errors inflate.
- Classical measurement error in **X**: OLS biased toward zero (attenuation).
- The attenuation factor is the **reliability** of $X$.
- Corrections: multi-item scales, SEM/factor models, IRT, 2SLS.
- The fix starts with better measurement, not better estimators.
